An integrated open population distance sampling approach for modelling age-structured populations

### 2023-12-18 Manuscript_Nilsen_Nater.qmd
Worked over Erlend's draft

Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005]. Over the last few decades, a tremendous progress have been made towards this end. This progress is partly driven the development and application of new field data collection methods and approaches, such as citizen science data (ref), camera trap data (ref) and the collection of environmental DNA data (ref). In addition, developments of novel statistical methods allow researchers to estimate abundance and demographic rates it situations where it was not feasible before (ref). Combined, these advance put us in a much better position to estimate state variables needed for population management (ref) or to estimate indexes that are relevant for large scale policy applications, e.g. such as Essential Biodiversity Variables [@Kissling2018].

Until recently, joint estimation of population dynamics and demography has relied mostly on data from marked individuals through application of open-population capture-mark-recapture models (ref). While such methods can provide valuable information for both ecological research and management, they are also in many cases costly to implement over large areas. On the other hand, data from monitoring of abundance trends over time typically are based on data from unmarked animals (ref). A common approach for such surveys is to structure data collection around the distance sampling (DS) method. DS have been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015]. One reason for their popularity might be that they does not require repeated visits to the same sites to estimate detection probability (ref). Rather, it is assumed that detection probability decrease as the objects are located longer away from the transect line (ref). This makes them particularily useful for implementation in participatory monitoring program, allowing stakeholders to take part in the data collection process.

The most common implementation of the DS model has been to use close-population formulations, that does not allow to formulate a process model that project abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates (ref). In recent years, DS approahces have been extended and also applied for estimating changes in abundance over time in open populations by embedding a hidden state model representing population dynamics [@moore2011; @sollmann2015]. Such approaches has greatly extended the application of DS approaches to study ecological dynamics in time and space. However, while these latter frameworks may allow to accurately quantify population changes but provide little information on the drivers of these changes, i.e. the underlying vital rates. In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no straight forward way to estimate demographic rates from such data. In cases where the age (and sex) of the observed individuals can be observed, this information can under some assumptions be used to inform about recruitment rates and survival probabilities (ref). For instance, in a integrated population model based on harvest statistics and observations of population age structure, both population abundance and demographic rates was estimated without even without additional data from marked individuals [@Nilsen2018].

Concurrent with the development of more sophisticated DS models, another group of models has emerged and gained popularity, not least for their ability to disentangle demographic processes underlying population dynamic: integrated population models [IPMs, @schaub2021]. Through joint analysis of multiple datasets, IPMs allow simultaneous estimation of population size and composition, as well as all vital rates that form part of an underlying typically age- or stage-structured population model. Since both DS models and IPMs estimate population size/density, a combination of the two frameworks has the potential to provide good estimates of both population- and demographic parameters and maximizing knowledge gained through transect surveys by augmenting them with other available data [e.g. @schmidt2020].

In this study, we will present a new IPM, where we integrate data from line transect distance sampling survey data and survival data from marked birds. The process model is a stage-structured matrix model, that projects population size from one time step to the next based on the underlying survival and recruitment rates. This is made possible by the fact that the line transect surveys provide data on population age structure which can be used to inform the estimation of demographic rates. In the IPM, we integrate these data with data from marked birds from which we estimate survival probabilities. The model is thus formally an IPM, as we jointly estimate parameters of interest based on several data sets (ref). In the paper, we first present the model and assess the robustness based on simulated data. In addition, we apply the model to real data collected from a willow ptarmigan (*Lagopus lagopus*) study in Central Norway. Because demographic rates are often affected by environmental covariates, we also illustrate how such covariates can be included in the modelling framework.

@@

Our open population integrated distance sampling model consists of two major components: a latent structured population model and a set of likelihoods for data originating from distance sampling surveys and auxiliary survival monitoring. In the example case, these auxiliary data come from a radio-telemetry study, but in principle other types of capture-recapture data can also be used.

@@

TBA

@@

The willow ptarmigan has a circumpolar distribution (Fuglei et al., 2020), and lives year-round in heterogeneous alpine and artic ecosystems. In Norway, there has been a long-term decline in the willow ptarmigan abundance spanning more than a century (ref), but the last few decades the trend in abundance has fluctuated both in time and space. In Scandinavia, willow ptarmigan is a valued game species (ref), and there has been several long-term research projects devoted to understanding how they respond to environmental variation and harvest management [@Israelsen2020; @Sandercock2011]. A key insight from across several study areas is the the annual recruitment rate (i.e. $R_t$ in our model, as outlined above) is highly variable, and is affected both by spring conditions [@Eriksen2023] and the abundance of small rodents as alternative prey for shared predators (i.e. the Alternative Prey Hypothesis ref). Adult survival show less inter-annual fluctuations [@Israelsen2020], although spatial (and potentially temporal) variation due to e.g. harvest management is evident when comparing across studies [@Israelsen2020].

Our case study was based on an ongoing long-term research project on willow ptarmigan in Lierne municipality in Central Norway (approximately 62.4 degrees north and 13.2 degrees east). The study area is located in a sub-alpine ecosystem, and the landscape is a mosaic of open heath and shrub vegetation (dominated by Ericacea, willow shrub *Salix spp*. and dwarf birch *Betula nana*), interspersed with bogs and forest patches (mainly birch *Betula spp*.). The climate is strongly seasonal, with snow typically covering the ground from October/November through April/May.

@@

1.  Data from a line transect survey program targeting willow ptarmigan operated as part of the natural resources management authorities

2.  Data from an individual based research project based on radio collared willow ptarmigan. The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2021 (and ongoing) and the individual based data spanning 2015-2021 (and ongoing).

Line transect survey data were collected in August each year, prior to the annual autumn harvest season, as part of the program "Hønsefuglportalen". Hønsefuglportalen is a national program for line transect surveys of tetraonid birds, and the effort is directed mainly to willow ptarmigan habitats. In our case study, we used data from the western part of Lierne municipality. Line transects are surveyed by educated volunteers, using pointing dogs to located the birds. When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds as well as the age (juvenile and adult) and sex of the birds are recorded. As the surveys are conducted in early August, juveniles can be distinguished from adults by their smaller body size. Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex. Since 2019, data has been collected through a mobile app tailormade for this project. Before 2019, field workers reported their data through a dedicated web portal. After data is collected, they go through several steps of quality control: first by local contacts and subsequently by personnel at NINA. The data are standardized based on the Darwin-Core standard (ref), and made publicly available as sampling-event data set published through GBIF [@Nilsen2023a]. For additional description of the data collection procedures, see [@Bowler2020]; [@Kvasnes2018][@Nilsen2023a].

The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2020. Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in [@Israelsen2020]. To prevent birds from flying off before the field personnel were close enough to capture them, a high-powered head lamp was used to dazzle the birds. After capture, birds were placed in an opaque bag to reduce stress. Before releasing the birds, they were fitted with a uniquely numbered leg ring (\~ 2.4g) and a Holohil RI-2BM or Holohil RI-2DM radio transmitter (\~ 14.1g). The radio transmitters had an expected battery lifetime of 24 months (RI-2BM) or 30 months (RI-2DM), and included a mortality circuit that was activated if a bird had been immobile for 12 hours. We monitored the birds throughout the year, by triangulation from the ground at least once a month for 10 months of the year (February -- November) by qualified field personnel. If a mortality signal was heard from the transmitter, we recovered it as soon as possible to determine cause of death. A number of birds dispersed out of the main study areas and was thus out of signal range for field personnel on the ground. To avoid loss of data, we conducted aerial triangulation using a helicopter or airplane three times a year (May, September and November) in the years 2016-2020.

### 2.4 Bayesian implementation

We implemented the model in a Bayesian framework using NIMBLE [@valpine2017] in R version 4.2.2 [@R2022]. The likelihood for line transect observation distances was set up using a custom half-normal distribution developed by Michael Scroggie as part of the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance). We used non-informative uniform priors (with biologically reasonable boundaries where possible) for all parameters. For each model fit, we ran x MCMC chains with NIMBLE's standard samples for xxxk iterations. xxxk thereof were discarded as burn-in prior to thining with a factor x, leaving us with xxxx posterior samples per chain (total of xxxxx samples per run).

@@

Based on simulated data, we assessed the models ability to estimate the underlying parameters. We run a total of 10 simulations, and the results are summarized in @fig-sim-detection and @fig-sim-demo. In general, the model was able to estimate both parameters of the detection model (Figure X) of the simulated data with little bias. This was true for all 10 simulations, and for underlying model parameters than span a wide variation in values. The full output from each simulation is available in the appendix.

![Model performance on simulated data. Upper panel depicts model parameters, and lower panels show derived estimated of effective strip width and mean detection probability. Dots represents true value, and density plot represent the posterior samples for each simulation](Figures/SimCheck_Detects_Temps.png){#fig-sim-detection}

![Posterior samples for the vital rates estimated based on simulated data. Upper panel show annual and seasonal survival probabilities, hyperparameters for the random effects model for recruitment (mean and sd), and lower panel show annual estimates of the recruitment rate. Dots represent true values, and density plots represents the posterior samples from each simulation](Figures/SimCheck_VRs_Temps.png){#fig-sim-demo}

@@

Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne as a case study to estimate abundance, vital rates and detection probabilities from a real-world data set. As is evident from Figure X, the population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021 (@fig-lierne-density). The increase was most distinct from 2016 and onward. Across years, mean survival probability for August - January ($S_1$) was estimated at `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$upper, 2)`), survival probability for February - July ($S_2$) was estimated as `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$upper, 2)`) (@fig-lierne-demo). Annual survival probability $S$ is given by the product of $S_1$ and $S_2$ was estimated at `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$upper, 2)`). Note that we assumed that $S$ was constant across all years in the study period from 2007 - 2021.

Recruitment ($R_t$) was allowed to vary across years (see model specification), and did as expected show large inter-annual variability. While the mean (baseline) recruitment $\mu_R$ was estimated as `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$Mean, 1)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$lower, 1)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$upper, 1)`). The estimated recruitment rate varied (as expected) among years: while the annual recruitment rate was estimated at `r round(max(filter(tidy_summary_stats, model_parameter =="R_year")$Mean), 1)` in year 2007, it was estimated at `r round(min(filter(tidy_summary_stats, model_parameter =="R_year")$Mean), 1)` in year 2012. Moreover, in contrast to the Alternative Prey Hypothesis (APH), our model did not find a clear link between the annual recruitment and the abundance of small rodents (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I. = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`).

@@

@@

-   In general, the model does a very good job in recreating the underlying parameters when fitted to simulated data where the true values are known. For the simulated data, a relative wide range of parameter values were allowed, and we did not detect any particular problems or biases with respect to this variation in the underlying parameters.
-   The fact that this formulation allow us to separately estimate the underlying demographic rates (survival and recruitment) is a clear improvement compared to previous open population DS models, as this allow us both to obtain estimates of the rates, but also to assess the effect of environmental variation directly on demographic rates and not just on population growth rate ($\lambda$) .
-   Nevertheless, we did not detect a clear effect of rodent abundance (indexed by the occurrence of rodents on the transect lines) on recruitment rate. This might be somewhat surprising, given that such a pattern has been reported repeatedly in the literature (see e.g. @Bowler2020). However, the data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small. \[A few sentences about fading rodent cycles, and perhaps a weakened link to ptarmigan dynamics\].
-   In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, creating a dependence between the demographic parameters. If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate. As a test of the phenomena, we also assessed how sensitive the model output was with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see **Methods**). In the version presented in the results section, we made the assumption that these birds were in fact juveniles. In an alternative scenario, we discarded all birds classified as unknowns, and rerun the model (Appendix). As expected the estimated population density was virtually unaffected by this decision, whereas the effect on the demographic parameters were somewhat more affected. Thus biases in the age ratio will affect the demographic rates, but not so much the estimated density. Compared to previous studies on ptarmigan, we could have expected a higher survival rate indicating that there is some misclassification in the data with respect to age specification.
-   The model is flexible, and can be extended with additional auxiliary data, such as separate data on recruitment rates that could improve the precision of the demographic rate estimates. It is also possible to extent the model with additional covariates, to gain a better understanding of how species respond to environmental variation through the effects on vital rates.

@@

EBN Lead - Drafted the model code; Lead - Wrote first version of ms; Lead; data collection Contributed - Modelling and analyses

CRN Lead - Updated and finalized model code; Lead - Modelling; Contributed - Wrote first version of ms

@@

>>>>

Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005, @williams2002]. Over the last few decades, tremendous progress has been made towards this end. This progress is partly driven the development and application of new field data collection methods and approaches, such as citizen science data (ref), camera trap data (ref) and the collection of environmental DNA data (ref). In addition, developments of novel statistical methods alongside decreases in computational costs now allow researchers to estimate abundance and demographic rates in situations where it was not feasible before (ref). Combined, these advances put us in a much better position for estimating quantities needed for population management [@williams2002] and indices relevant for large scale policy applications, e.g. Essential Biodiversity Variables [@Kissling2018].

Until recently, joint estimation of population dynamics and demography has relied mostly on data from marked individuals through application of open-population capture-mark-recapture models [@schaub2021]. While such methods can provide valuable information for both ecological research and management, collecting the necessary data is typically costly and logistically challenging to implement over large areas. Monitoring focusing on abundance trends over larger areas, on the other hand, are typically based on data from unmarked animals (ref). A common approach for such surveys is to structure data collection around the distance sampling (DS) method. DS has been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015]. One reason for the method's popularity is that it does not require repeated visits to the same sites for estimating detection probability (ref). Rather, it is assumed that detection probability decreases the further away from the transect line animals are located (ref). This makes DS particularly useful for implementation in participatory monitoring programs, allowing stakeholders to take part in the data collection process. <!-- #It does not become evident why the fact that detection is assumed to decrease with distance makes a survey design more suitable for participatory monitoring. May need some more explanation. -->

The most common implementations of DS models have long used close-population formulations, and thus do not allow formulating a process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates (ref). In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015]. This has greatly extended the potential of DS approaches for studying ecological dynamics across time and space. However, while these latter frameworks may allow to accurately quantify population changes, they typically provide little information on the drivers of these changes, i.e. the underlying vital rates. In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no straightforward way to estimate demographic rates from such data. On the contrary, if age (and sex) of detected individuals can be determined, this information can be used to provide information on recruitment rates and survival probabilities (ref). @Nilsen2018, for example, used a model based on harvest statistics and observations of population age structure to estimate population abundance and demographic rates without the need for any additional data from marked individuals. <!-- #I have removed the mention of integrated population models from this sentence since those only get specifically introduced in the next paragraph -->

Concurrent with the development of more sophisticated DS models, another group of models has emerged and rapidly gained popularity, not least for their ability to disentangle demographic processes underlying population dynamic: integrated population models [IPMs, @schaub2021]. Through joint analysis of multiple datasets, IPMs allow simultaneous estimation of population size and composition, as well as all vital rates that form part of an underlying age- or stage-structured population model. Since both DS models and IPMs estimate population size/density, a combination of the two frameworks has the potential to provide good estimates of both population- and demographic parameters by maximizing knowledge gained from transect surveys by augmenting them with other available data [e.g. @schmidt2020].

In this study, we present a new IPM (IDSM, integrated distance sampling model) which integrates data from line transect distance sampling survey data and survival data from marked birds. The model's core is a stage-structured matrix population model that projects population size from one time step to the next based on underlying survival and recruitment rates. <!-- # I suggest removing the following two sentences as they mostly re-state information we have already provided: "This is made possible by the fact that the line transect surveys provide data on population age structure which can be used to inform the estimation of demographic rates. In the IPM, we integrate these data with data from marked birds from which we estimate survival probabilities. The model is thus formally an IPM, as we jointly estimate parameters of interest based on several data sets (ref)." --> Below, we first present the model and assess its robustness and performance through application to simulated data. We then apply the model to real data collected from a willow ptarmigan (*Lagopus lagopus*) study in Central Norway. Because demographic rates are often affected by environmental covariates (e.g. rodent abundance in the case of willow ptarmigans), we also illustrate how such covariates can be included in the modelling framework.

@@

Our open population integrated distance sampling model (IDSM) consists of two major components: a latent structured population model and a set of likelihoods for data originating from distance sampling surveys and auxiliary survival monitoring. In the example case, these auxiliary data come from a radio-telemetry study, but in principle other types of capture-recapture data can also be used.

@@

We assessed the model's overall performance and ability to estimate abundance, demographic rates, and detection parameters without bias by testing it on simulated data. We generated a total of 10 simulated datasets in five steps. First, we simulated 15-year time-series of survival and recruitment rates from biologically plausible values for averages of and -- in the case of recruitment -- among-year variation in demographic rates (survival was held constant across years). Second, we used the yearly demographic rate and realistic initial population densities to simulate stochastic population dynamics in 50 distinct sites. Third, we simulated the grouping of individuals in each site by first determining the expected number of groups in a site (based on the average group size of 5.6 from our ptarmigan data) and then distributing individuals among groups via multinomial trials. Fourth, we assigned a distance from transect line to each group and simulated the line transect survey in all 50 sites across 15 years. Finally, we simulated 5-year time-series of radio-telemetry data (= survival from one year to the next) for a subset of individuals (30 on average) using the simulated survival probabilities for each relevant year. We then fit the IDSM to each of the 10 simulated datasets three times, using distinct seeds for both simulating initial values and initiating and running the MCMC. Model implementation for simulated data tests was largely identical to that for real data and is described in detail below.

@@

The willow ptarmigan has a circumpolar distribution (Fuglei et al., 2020), and lives year-round in heterogeneous alpine and artic ecosystems. In Norway, there has been a long-term decline in the willow ptarmigan abundance across more than a century (ref), but in the last few decades abundance trends has fluctuated both in time and space. In Scandinavia, willow ptarmigan is a valued game species (ref), and there have been several long-term research projects devoted to understanding how they respond to environmental variation and harvest management [@Israelsen2020; @Sandercock2011]. A key insight from across several study areas is the the annual recruitment rate (i.e. $R_t$ in our model, as outlined above) is highly variable, and is affected both by spring conditions [@Eriksen2023] and the abundance of small rodents, which constitute alternative prey for common predators (i.e. the Alternative Prey Hypothesis ref). Adult survival show less inter-annual fluctuations [@Israelsen2020], although spatial (and potentially temporal) variation due to e.g. harvest management is evident when comparing across studies [@Israelsen2020].

Our case study was based on an ongoing long-term research project on willow ptarmigan in Lierne municipality in Central Norway (approximately 62.4 degrees north and 13.2 degrees east). The study area is located in a sub-alpine ecosystem, and the landscape is a mosaic of open heath and shrub vegetation (dominated by Ericacea, willow shrub *Salix spp*., and dwarf birch *Betula nana*), interspersed with bogs and forest patches (mainly birch *Betula spp*.). The climate is strongly seasonal, with snow typically covering the ground from October/November through April/May.

@@

1.  Data from a line transect survey program targeting willow ptarmigan operated under the natural resources management authorities (2007-2021, ongoing)

2.  Data from an individual-based monitoring programme based on radio collared willow ptarmigan (2015-2021, ongoing)

Line transect survey data were collected in August each year, prior to the annual autumn harvest season, as part of the program "Hønsefuglportalen". Hønsefuglportalen is a national program for line transect surveys of tetraonid birds, and the effort is directed mainly towards willow ptarmigan habitats. In our case study, we used data from the western part of Lierne municipality. Line transects are surveyed by trained volunteers that use pointing dogs to locate the birds. When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds in the group, as well as the age (juvenile or adult) and sex of the birds are recorded. As the surveys are conducted in early August, juveniles can be distinguished from adults by their smaller body size. <!-- maybe also mention how males and females are distinguished? Also, is it just body size for the juveniles or also plumage? --> Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex. Since 2019, data has been collected through a mobile app tailormade for this project <!-- Add name + maybe link to download in app store? --> . Before 2019, field workers reported their data through a dedicated web portal. After data are collected and reported, they undergo several steps of quality control: first by local contacts and subsequently by personnel at the Norwegian Institute for Nature Research (NINA). The data are then standardized based on the Darwin-Core standard (ref), and made publicly available as a sampling-event data set published through GBIF [@Nilsen2023a]. For additional description of the data collection procedures, see [@Bowler2020]; [@Kvasnes2018][@Nilsen2023a].

The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2020. <!-- Above you stated that radio telemetry data collection is until 2021 and ongoing. Until what year was it? --> Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in [@Israelsen2020]. To prevent birds from flying off before the field personnel were close enough to capture them, a high-powered head lamp was used to dazzle the birds. After capture, birds were placed in an opaque bag to reduce stress. They were then fitted with a uniquely numbered leg ring (\~ 2.4g) and a Holohil RI-2BM or Holohil RI-2DM radio transmitter (\~ 14.1g) and subsequently released. The radio transmitters had an expected battery lifetime of 24 months (RI-2BM) or 30 months (RI-2DM), and included a mortality circuit that was activated if a bird had been immobile for 12 hours. We monitored the birds throughout most of the year by triangulation from the ground at least once a month for 10 months of the year (February -- November) by qualified field personnel. If a mortality signal was emitted from a transmitter, we attempted to recover it as soon as possible to determine cause of death. A number of birds dispersed out of the main study areas and was thus out of signal range for field personnel on the ground. To avoid loss of data, we conducted aerial triangulation using a helicopter or airplane three times a year (May, September and November) in the years 2016-2020.

### 2.4 Bayesian model implementation

We implemented the model in a Bayesian framework using NIMBLE version 1.0.1 [@valpine2017] in R version 4.3.1 [@R2023]. The likelihood for line transect observation distances was set up using a custom half-normal distribution developed by Michael Scroggie as part of the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance). We used non-informative uniform priors (with biologically reasonable boundaries where possible) for all parameters. We assumed constant survival and time-varying recruitment rate in models fit to both simulated and real data.

For the model fits to simulated and real data we ran 3 and 4 MCMC chains with NIMBLE's standard samples for 500k and 100k iterations, respectively. 300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), repectively.

@@

Posterior distributions for parameters estimated in three model fits to each of 10 simulated data sets are shown in @fig-sim-detection and @fig-sim-demo. Overall, the IDSM was able to corretly estimate both detection parameters and demographic rate parameters from all 10 simulated datasets without any systematic bias. The replicate runs for each dataset resulted in very similar posterior distributions, demonstrating that the models converged towards the same posterior distributions irrespective of starting values. Estimates of population sizes / densities were also adequate, and are presented in the Supplementary Materials (see folder SimCheck_byDataSet).

![Detection parameters estimated from 10 distinct sets of simulated data (= colors). Upper panel depicts model parameters, and lower panels show derived estimated of effective strip width and mean detection probability. Dots represents true values (black for global values, colored for dataset-specific values), and density plot represent the posterior distributions for each model run. ](Figures/SimCheck_Detects_Temps.png){#fig-sim-detection}

![Demographic rate parameters estimated from 10 distinct sets of simulated data (= colors). Upper panel show annual and seasonal survival probabilities, hyperparameters for the random effects model for recruitment (mean and sd), and lower panel show annual estimates of the recruitment rate. Dots represent true values (black for global values, colored for dataset-specific values), and density curves represent the posterior distributions from each model run. ](Figures/SimCheck_VRs_Temps.png){#fig-sim-demo}

@@

Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne as a case study to estimate abundance, vital rates and detection probabilities from a real-world data set. Ptarmigan population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021 (@fig-lierne-density). The increase was most distinct from 2016 and onward.

@@

Average survival probability for August - January ($S_1$) was estimated at `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$upper, 2)`) while average survival probability for February - July ($S_2$) was estimated as `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$upper, 2)`) (@fig-lierne-demo A). Annual survival probability $S$, given by the product of $S_1$ and $S_2$, was estimated at `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$upper, 2)`), @fig-lierne-demo A).

Recruitment ($R_t$) was allowed to vary across years (see model specification), and estimates displayed large inter-annual variability (@fig-lierne-demo C, Figure S1). While the mean (baseline) recruitment $\mu_R$ was estimated as `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$Mean, 1)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$lower, 1)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$upper, 1)`) the yearly recruitment rates ranged from `r round(min(filter(tidy_summary_stats, model_parameter =="R_year")$Mean), 1)` in year 2012 to `r round(max(filter(tidy_summary_stats, model_parameter =="R_year")$Mean), 1)` in year 2007.

Given the available data, the IDSM was not able to estimate a clear effect of small rodent abundance on ptarmigan recruitment (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I. = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`).

@@

-   In general, the model does a very good job in recreating the underlying parameters when fitted to simulated data where the true values are known. For the simulated data, a relative wide range of parameter values were allowed, and we did not detect any particular problems or biases with respect to this variation in the underlying parameters.
-   Both simulated data tests and real data application have shown that mixing is suboptimal for the average recruitment rate parameter (Mu.R). Future work should perhaps look into optimizing sampling? (For biological interpretation that is not a big problem as the yearly recruitment rate estimates are fine... but have to keep in mind when interpreting) <!-- I leave it to you Erlend to decide where, whether, and how much you want to go into this one -->
-   The fact that this formulation allow us to separately estimate the underlying demographic rates (survival and recruitment) is a clear improvement compared to previous open population DS models, as this allow us both to obtain estimates of the rates, but also to assess the effect of environmental variation directly on demographic rates and not just on population growth rate ($\lambda$) .
-   Nevertheless, we did not detect a clear effect of rodent abundance (indexed by the occurrence of rodents on the transect lines) on recruitment rate. This might be somewhat surprising, given that such a pattern has been reported repeatedly in the literature (see e.g. @Bowler2020). However, the data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small. \[A few sentences about fading rodent cycles, and perhaps a weakened link to ptarmigan dynamics\]. \[Also: a time-series of 15 years is still relatively short, so it is to be expected that we have low statistical power to detect temporal covariate effects --\> more years of data, or space-for-time substitution, will help)
-   In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, creating a dependence between the demographic parameters. If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate. As a test of the phenomena, we also assessed how sensitive the model output was with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see **Methods**). In the version presented in the results section, we made the assumption that these birds were in fact juveniles. In an alternative scenario, we discarded all birds classified as unknowns, and rerun the model (Appendix). As expected the estimated population density was virtually unaffected by this decision, whereas the effect on the demographic parameters were somewhat more affected. Thus biases in the age ratio will affect the demographic rates, but not so much the estimated density. Compared to previous studies on ptarmigan, we could have expected a higher survival rate indicating that there is some misclassification in the data with respect to age specification.
-   The model is flexible, and can be extended with additional auxiliary data, such as separate data on recruitment rates that could improve the precision of the demographic rate estimates. Another one might be data that allows to distinguish between true and apparent survival<!-- or is that accounted for in the telemetry data, i.e. are the individuals that left hte study area censored? -->. It is also possible to extent the model with additional covariates, to gain a better understanding of how species respond to environmental variation through the effects on vital rates. \[See my point above on limited statistical power with 15 years. Maybe could be good to link to the space-for-time substitution idea here and "tease" the extension to multi-area since the monitoring programme, at least for line transects, covers many more areas).

@@

EBN Lead - Drafted the model code; Lead - Wrote first version of ms; Lead - Data collection; Contributed - Modelling and analyses

CRN Lead - Updated, further developed, and finalized model code; Lead - Modelling and analyses; Contributed - Wrote first version of ms

@@

\[Should we thank James for discussions etc.?\]
